---
title: "Exploration of bibliometrix R package"
author: Romain Frelat
date: today
format: 
  html:
    toc: true
execute:
  echo: true
  warning: false
---


The objective is to explore the functionalities of the [bibliometrix R package](https://github.com/massimoaria/bibliometrix).   

The example case study is based on the references cited by projects financed under the [FRB-MTE-OFB projects](https://www.fondationbiodiversite.fr/la-frb-en-action/programmes-et-projets/impacts-sur-la-biodiversite-terrestre-dans-lanthropocene/). The references were cleaned and completed in the scripts [01_clean_references.R](https://codeberg.org/frbcesab/Biblioscape/src/branch/main/analysis/01_clean_references.R) and [02_fetch_openalex.R](https://codeberg.org/frbcesab/Biblioscape/src/branch/main/analysis/02_fetch_openalex.R).  

The bibliometrix package has a nice [Shiny interface](https://www.bibliometrix.org/biblioshiny/) that can be accessed locally with:
```{r}
#| eval: false
bibliometrix::biblioshiny()
```

The [openalex data](https://codeberg.org/frbcesab/Biblioscape/src/branch/main/data/derived-data/mte_references_oa.rdata) in can be loaded in the Shinyapp in the section Data > Import raw files > OpenAlexAPI.   

But here the objective is to use and explore the functions of [`bibliometrix` package](https://github.com/massimoaria/bibliometrix).

## Data loading and converting

The data loading takes few minutes to run.

```{r}
library(bibliometrix)
oadata <- here::here("data", "derived-data", "mte_references_oa.rdata")

M <- bibliometrix::convert2df(
  file = oadata,
  dbsource = "openalex_api",
  format = "api"
)
```

<!-- 
Removed  97 duplicated documents
Warning message:
3 record(s) have a suspicious authorship list
(likely OpenAlex source corruption: cited-reference authors absorbed into the author field).
Flagged in M$AU_FLAG; inspect row(s) 259, 936, 1134 before running collaboration networks. 
-->

We can check the quality of the bibliometric data.

```{r}
com <- missingData(M)
print(com$mandatoryTags)
```

## Simple bibliometric Analysis

A recommended first step is a descriptive analysis of the bibliographic data with the function [`biblioAnalysis`](https://cran.r-project.org/web/packages/bibliometrix/refman/bibliometrix.html#biblioAnalysis). It calculates:

- Annual scientific production
- Most productive authors
- Most cited manuscripts
- Most productive countries
- Total citations per country
- Most relevant sources (journals)
- Most frequent keywords

```{r}
results <- biblioAnalysis(M, sep = ";")
S <- summary(object = results, k = 10, pause = FALSE)
plot(x = results, k = 10, pause = FALSE)
```


### Local citations

Local citations measure how many times an author (or a document) included in the collection have been cited within the collection.

```{r}
CR <- localCitations(M, sep = ";")
CR$Authors[1:10, ]
CR$Papers[1:10, ]
```

### Zoom on authors

```{r}
DF <- dominance(results, k = 10)
DF
```

```{r}
# Top-Authors’ Productivity over the Time
topAU <- authorProdOverTime(M, k = 10, graph = TRUE)
```


<!-- 
### Life cycle analysis
data <- M %>% group_by(PY) %>% count()
LC <- lifeCycle(data, forecast_years = 20, plot = TRUE, verbose = FALSE) 
-->



## Bibliographic network matrices

### Co-Citation Network

```{r}
NetMatrix <- biblioNetwork(
  M,
  analysis = "co-citation",
  network = "references",
  n = 30,
  sep = ";"
)

# Plot the network
net <- networkPlot(
  NetMatrix,
  Title = "Co-Citation Network",
  type = "fruchterman",
  size = T,
  remove.multiple = FALSE,
  labelsize = 0.7,
  edgesize = 5
)
```

### Keyword co-occurrences


```{r}
# Create keyword co-occurrences network
NetMatrix <- biblioNetwork(
  M,
  analysis = "co-occurrences",
  network = "keywords",
  sep = ";"
)

# Plot the network
net <- networkPlot(
  NetMatrix,
  normalize = "association",
  weighted = T,
  n = 30,
  Title = "Keyword Co-occurrences",
  type = "fruchterman",
  size = T,
  edgesize = 5,
  labelsize = 0.7
)
```


## Thematic map

### Co-Word Analysis

```{r}
CS <- conceptualStructure(
  M,
  field = "ID",
  method = "MCA",
  minDegree = 10,
  clust = 6,
  stemming = FALSE,
  labelsize = 15,
  documents = 20,
  graph = FALSE
)
plot(CS$graph_terms)
# plot(CS$graph_dendogram)
```

### Thematic map

```{r}
thematicMapResults <- thematicMap(
  M,
  field = "DE",
  n = 250,
  minfreq = 5,
  stemming = FALSE,
  size = 0.3,
  n.labels = 3,
  repel = TRUE,
  cluster = "louvain"
)
plot(thematicMapResults$map)

plot(thematicMapResults$net$graph)
```