---
title: "Exploration of group memberships"
author: Romain Frelat
date: today
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
---


The objective is to explore the relation among members of [FRB-MTE-OFB projects](https://www.fondationbiodiversite.fr/la-frb-en-action/programmes-et-projets/impacts-sur-la-biodiversite-terrestre-dans-lanthropocene/). The raw data is not available on Github to safeguard privacy.


## General overview

```{r}
devtools::load_all()
library(igraph)
library(visNetwork)

# load membership list
mem <- read.csv(
  here::here("data", "derived-data", "mte_members.csv")
)

# load informations on projects
meta <- readxl::read_xlsx(
  here::here("data", "raw-data", "mte", "Projets_FRB-MTECT-OFB.xlsx")
)

# create a last+first name column
mem$name <- paste(mem$lastname, mem$firstname)

# number of members per group and position
table(mem$group, mem$function_in_group)

# number of members per type of group
n_mem <- data.frame(table(mem$group))
n_mem$Type <- meta$Type[match(n_mem$Var1, meta$Acronyme)]
tapply(n_mem$Freq, n_mem$Type, summary)

# list all group per name
infoV <- tapply(mem$group, mem$name, paste, collapse = "<br>")

nproj <- table(table(mem$name))
```

Among the ```r nrow(meta)``` projects of the programs, there are ```r length(unique(mem$name))``` different researchers. Most of the researcher are member of a single project (N=```r nproj["1"]```), but ```r nproj["2"]``` researchers are attached to 2 projects; and ```r nproj["3"]``` researchers works on 3 projects.  



## Project network

Based on the number of shared members, we can build a network of projects.

```{r}
group_mem <- table(mem$group, mem$name)
member_overlap <- tcrossprod(group_mem)
# or Jaccard to account for different group size
# proxy::simil(group_mem, method = "Jaccard", diag = TRUE, upper = TRUE)

diag(member_overlap) <- 0

netMEM <- igraph::graph_from_adjacency_matrix(
  member_overlap,
  mode = "undirected",
  weighted = TRUE
)

modMEM <- cluster_walktrap(netMEM)
```

The network has ```r vcount(netMEM)``` nodes, ```r ecount(netMEM)``` edges, a connectance of ```r round(edge_density(netMEM, loops = FALSE),2)``` and a modularity of ```r round(modularity(modMEM),2)```


### Project network with modules
```{r}
modcol <- rainbow(max(membership(modMEM)))[membership(modMEM)]
vtyp <- as.factor(meta$Type[match(V(netMEM)$name, meta$Acronyme)])
vyr <- meta$Year[match(V(netMEM)$name, meta$Acronyme)]

V(netMEM)$color <- modcol
E(netMEM)$width <- E(netMEM)$weight
V(netMEM)$title <- paste(V(netMEM)$name, vtyp, vyr, sep = "<br>")
visNetwork::visIgraph(netMEM, randomSeed = 54, layout = "layout_nicely")
# layout_nicely, layout_with_fr, layout_with_kk
```

### Project network with year and type

```{r}
# color by typexnod
typcol <- rainbow(nlevels(vtyp))[vtyp]
V(netMEM)$color <- typcol

# x-axis by year
xnod <- (vyr - min(vyr)) / (max(vyr) - min(vyr)) * 2 - 1

# take the membership as y-axis
ymod <- ifelse(membership(modMEM) > 5, 6, membership(modMEM))
ynod <- ymod / max(ymod)
# layout_nicely(netMEM)[, 2]
# sqrt(membership(modMEM)) / max(sqrt(membership(modMEM)))
# or take the first MDS dimension
# ynod <- layout_with_mds(netMEM)[, 1]

# coordinates
coo <- cbind(jitter(xnod, 1.5), jitter(ynod, 3))

# plot(netMEM, layout = coo)
laby <- sort(unique(vyr))
seqy <- (laby - min(laby)) / (max(laby) - min(laby)) * 2 - 1

visNetwork::visIgraph(
  netMEM,
  layout = 'layout.norm',
  layoutMatrix = coo,
  randomSeed = 54,
  type = "full"
)
```

### Centrality

There are a lot of different measures of centrality; here we use betweenness (number of shortest paths going through a vertex) and the Google page rank
```{r}
betweenMEM <- betweenness(netMEM)
degreeMEM <- degree(netMEM)
prMEM <- page_rank(netMEM)$vector
par(mar = c(4, 4, 3, 1))
plot(
  betweenMEM,
  prMEM,
  xlab = "Betweenness",
  ylab = "Page rank",
  main = "Project centrality",
  type <- "n"
)

text(betweenMEM, prMEM, V(netMEM)$name, xpd = NA)
```


## Researcher network

Based on project membership, we can build two member networks:

a. where all group members are linked among them
b. where group members are linked only to the PI+postdoc

```{r}
pw <- c()
pw_pi <- c()
for (i in sort(unique(mem$group))) {
  ni <- mem$name[mem$group == i]
  pwi <- expand.grid(ni, ni) #or combn(ni, 2) for undirected
  combn(ni, 2)
  #remove identity / loop
  pwi <- pwi[pwi$Var1 != pwi$Var2, ]
  pw <- rbind(pw, pwi)

  # considering the central role of PI
  pi <- mem$name[
    mem$function_in_group %in% c("PI", "Post-doc") & mem$group == i
  ]
  pwi2 <- expand.grid(ni, pi)
  pwi2 <- pwi2[as.character(pwi2$Var1) != as.character(pwi2$Var2), ]
  pw_pi <- rbind(pw_pi, pwi2)
}

# add weight based on how many times
netRES <- igraph::graph_from_data_frame(pw, directed = FALSE)
modRES <- cluster_walktrap(netRES)
# plot(
#   netRES,
#   layout = layout_with_mds(netRES),
#   vertex.label = NA
# )

netRES_PI <- igraph::graph_from_data_frame(pw_pi, directed = FALSE)
modRES_PI <- cluster_walktrap(netRES_PI)
# plot(
#   netRES_PI,
#   layout = layout_with_mds(netRES_PI),
#   vertex.label = NA
# )
```




### A. all members

The network linking all group members between them has ```r vcount(netRES)``` nodes, ```r ecount(netRES)``` edges, a connectance of ```r round(edge_density(netRES, loops = FALSE),2)``` and a modularity of ```r round(modularity(modRES),2)```. 

```{r}
V(netRES)$title <- paste(
  V(netRES)$name,
  infoV[match(V(netRES)$name, names(infoV))],
  sep = "<br>"
)

modcol <- rainbow(max(membership(modRES)))[membership(modRES)]
V(netRES)$color <- modcol
visNetwork::visIgraph(netRES)
```

### B. PI connected

The network linking all group members to the PI+Postdoc has ```r vcount(netRES_PI)``` nodes, ```r ecount(netRES_PI)``` edges, a connectance of ```r round(edge_density(netRES_PI, loops = FALSE),2)``` and a modularity of ```r round(modularity(modRES_PI),2)```. 

```{r}
V(netRES_PI)$title <- paste(
  V(netRES_PI)$name,
  infoV[match(V(netRES_PI)$name, names(infoV))],
  sep = "<br>"
)

modcol <- rainbow(max(membership(modRES_PI)))[membership(modRES_PI)]
V(netRES_PI)$color <- modcol
visNetwork::visIgraph(netRES_PI)
```


## Bipartite network


```{r}
netBI <- graph_from_data_frame(
  mem[, c("name", "group")]
)
V(netBI)$type <- V(netBI)$name %in% meta$Acronyme
V(netBI)$color <- ifelse(V(netBI)$type, "blue", "red")
# V(netBI)$shape <- ifelse(V(netBI)$type, "square", "circle")
# V(netBI)$size <- 1
modBI <- cluster_walktrap(netBI)

# plot(netBI, layout = layout_as_bipartite(netBI))
visNetwork::visIgraph(netBI)
```

The bipartite network has ```r vcount(netBI)``` nodes, ```r ecount(netBI)``` edges, a connectance of ```r round(edge_density(netBI, loops = FALSE),4)``` and a modularity of ```r round(modularity(modBI),2)```. 