---
title: "Exploration of project citations"
author: Romain Frelat
date: today
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
---


The objective is to explore the associations among [FRB-MTE-OFB projects](https://www.fondationbiodiversite.fr/la-frb-en-action/programmes-et-projets/impacts-sur-la-biodiversite-terrestre-dans-lanthropocene/) based on shared references. 

Based on references, we:  

- quantify the overlap among projects  
- cluster projects based on shared references 
- identify papers that link multiple projects (*hubs*)


This document is mostly an exploration of how network analysis methodology.  

*The raw data is not available on Github to safeguard privacy.*


## General overview

```{r}
devtools::load_all()
library(igraph) # network analysis
library(visNetwork) # interactive network
library(plotly) # interactive plot

# load reference list
ref <- read.csv(
  here::here("data", "derived-data", "mte_references_completed.csv")
)
ref$Project <- firstup(ref$Project)
# remove references with no DOI
ref <- ref[!is.na(ref$DOI), ]

# load informations on projects
meta <- readxl::read_xlsx(
  here::here("data", "raw-data", "mte", "Projets_FRB-MTECT-OFB.xlsx")
)

ndoi <- table(table(ref$DOI))

load(
  here::here("data", "derived-data", "mte_references_oa_shinycomplete.RData")
)
# M <- readRDS(here::here("data", "derived-data", "mte_bibliometrix.rds"))
# table(ref$DOI %in% M$DI)
ref$Short <- M$SR[match(ref$DOI, M$DI)]
ref$FAut <- M$AU_CORR[match(ref$DOI, M$DI)]
```

```{r}
# table(ref$Year, ref$Project)
table(ref$Project, ref$Relation)

# between 11 and 116 references per project
# summary(as.numeric(table(ref$Project)))
n_ref <- data.frame(table(ref$Project))
n_ref$Type <- meta$Type[match(n_ref$Var1, meta$Acronyme)]
tapply(n_ref$Freq, n_ref$Type, summary)
```


Each project cited between ```r min(n_ref$Freq)``` and ```r max(n_ref$Freq)``` references in their proposal. These references comes from the main text (`Proposal`), the top articles per member (`Top5`) or listed as `Output` in the final report. There are strong differences in the proposal format depending on the project type, `synthese` being the most complete proposal compare to `synergie` and `revue systematique`.  

In total, there are ```r nrow(ref)``` references listed by the ```r nrow(n_ref)``` projects. Most publications are cited by a single project (N=```r ndoi["1"]```), but ```r sum(ndoi[-1])``` publications are cited by at least two projects.


## Bipartite network

We can build a bi-partite network with references connecting projects.

```{r}
netBI <- graph_from_data_frame(
  ref[, c("DOI", "Project")]
)
V(netBI)$type <- V(netBI)$name %in% ref$Project
# V(netBI)$size <- 1
modBI <- cluster_walktrap(netBI)
```

The bipartite network has ```r vcount(netBI)``` nodes, ```r ecount(netBI)``` edges, a connectance of ```r round(edge_density(netBI, loops = FALSE),3)``` and a modularity of ```r round(modularity(modBI),2)```. 

```{r}
V(netBI)$color <- ifelse(V(netBI)$type, "red", "blue")
V(netBI)$shape <- ifelse(V(netBI)$type, "square", "dot")

# plot(netBI, layout = layout_as_bipartite(netBI))
visNetwork::visIgraph(netBI)
```


To analyse the bipartite network, it is best to split it into two networks:
- a project network to explore the links among projects
- a reference network to explore the centrality of references


```{r}
splitBI <- bipartite_projection(netBI)
netREF <- splitBI$proj1
netPR <- splitBI$proj2
```

### Project network

Based on the number of shared references, we can build a network of projects.

```{r}
# group_mem <- table(mem$group, mem$name)
# member_overlap <- tcrossprod(group_mem)
# diag(member_overlap) <- 0
# netMEM <- igraph::graph_from_adjacency_matrix(
#   member_overlap,
#   mode = "undirected",
#   weighted = TRUE
# )
modPR <- cluster_walktrap(netPR)
cluPR <- membership(modPR)
nclu <- table(cluPR)
cclu <- colorspace::qualitative_hcl(n = length(nclu), palette = "Dark 3")
cclu[nclu == 1] <- "#808080" # grey
colPR <- cclu[cluPR]
simclu <- ifelse(cluPR <= (sum(nclu > 1)), cluPR, "isolated")
```



The network has ```r vcount(netPR)``` nodes (one node = one project) and ```r ecount(netPR)``` edges (one edge = shared reference between two projects). The connectance is ```r round(edge_density(netPR, loops = FALSE),3)``` and the modularity ```r round(modularity(modPR),2)```


#### Modules
```{r}
V(netPR)$type <- as.factor(meta$Type[match(V(netPR)$name, meta$Acronyme)])
V(netPR)$shape <- c("triangle", "square", "dot")[V(netPR)$type]
V(netPR)$year <- meta$Year[match(V(netPR)$name, meta$Acronyme)]
V(netPR)$group <- simclu
V(netPR)$color <- colPR
E(netPR)$width <- E(netPR)$weight
V(netPR)$title <- paste(
  V(netPR)$name,
  V(netPR)$type,
  V(netPR)$year,
  sep = "<br>"
)
visNetwork::visIgraph(netPR, randomSeed = 54, layout = "layout_nicely") |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "group", multiple = T)
  ) |>
  visLegend(
    addNodes = list(
      list(
        label = "Revue",
        shape = "triangle",
        color = "grey"
      ),
      list(
        label = "Synergie",
        shape = "square",
        color = "grey"
      ),
      list(
        label = "Synthese",
        shape = "dot",
        color = "grey"
      )
    ),
    useGroups = FALSE
  ) |>
  visPhysics(stabilization = TRUE)
# layout_nicely, layout_with_fr, layout_with_kk
```

```{r}
# identify cliques
for (i in sort(unique(simclu))) {
  print(paste(
    i,
    paste(V(netPR)$name[simclu == i], collapse = ", "),
    sep = ": "
  ))
}
```


## Reference network

Based on the reference network, we can identify the centrality of shared references. 


```{r}
# show the weight
E(netREF)$width <- E(netREF)$weight

V(netREF)$size <- log(degree(netREF))

V(netREF)$label <- ref$Short[match(V(netREF)$name, ref$DOI)]
V(netREF)$year <- ref$Year[match(V(netREF)$name, ref$DOI)]
V(netREF)$faut <- ref$FAut[match(V(netREF)$name, ref$DOI)]
V(netREF)$fulltitle <- ref$Title[match(V(netREF)$name, ref$DOI)]
V(netREF)$shorttitle <- substr(V(netREF)$fulltitle, 1, 25)

V(netREF)$title <- paste(
  V(netREF)$name,
  V(netREF)$year,
  V(netREF)$faut,
  V(netREF)$shorttitle,
  sep = "<br>"
)

# plot(netREF)
# visNetwork::visIgraph(netREF, randomSeed = 54, layout = "layout_nicely") |>
#   visOptions(
#     highlightNearest = TRUE,
#   )
```

#### Centrality

There are a lot of different measures of centrality. Here we use betweenness (number of shortest paths going through a vertex) and the Google page rank.


```{r}
centR <- data.frame(
  "betweenness" = betweenness(netREF),
  "degree" = degree(netREF),
  "pagerank" = page_rank(netREF)$vector,
  "lab" = V(netREF)$title
)

plot_ly(centR) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Researcher centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
info <- apply(centR[, -4], 2, round, 3)
score <- apply(scale(sqrt(info)), 1, mean)
info <- as.data.frame(info)
info$first_aut <- V(netREF)$faut
info$year <- V(netREF)$year
info$short_title <- V(netREF)$shorttitle
DT::datatable(info[order(score, decreasing = TRUE)[1:50], ])
```