---
title: "Exploration of co-citation networks"
author: Romain Frelat
date: today
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
---


The objective of this document is to explore the co-citations among the [FRB-MTE-OFB projects](https://www.fondationbiodiversite.fr/la-frb-en-action/programmes-et-projets/impacts-sur-la-biodiversite-terrestre-dans-lanthropocene/). The references were cleaned and completed in the scripts [01_clean_references.R](https://github.com/FRBCesab/biblioscape/blob/main/analysis/01_clean_references.R) and [02_fetch_openalex.R](https://github.com/FRBCesab/biblioscape/blob/main/analysis/02_fetch_openalex.R).   

Estimate the distance among papers cited by projects using co-citations.
The questions are:
- how wide/narrow are the research projects?
- how similar/interconnected are the projects? can we make group of projects based on their citations?
- which are key work that are cited by many projects?

Nodes are all papers cited by the projects, and cited papers that are cited by at least two papers.

> The Co-Citation Network maps the intellectual structure of a research field by analyzing patterns in how documents cite the same references. Two references are co-cited when they are both cited by the same document. The more frequently two references are co-cited, the stronger their intellectual relationship. This technique, introduced by Small (1973), is one of the foundational methods of bibliometric analysis.
>
> Co-citation analysis rests on two complementary approaches:  
> - Co-citation (Small, 1973): Measures how often two references are cited together by subsequent publications. Highly co-cited references are perceived as intellectually related by the citing community. This method identifies the knowledge base of a field.
> - Bibliographic coupling (Kessler, 1963): Measures the overlap in the reference lists of two documents. Documents that share many cited references are likely working on similar topics. This method identifies current research fronts.
>
> Both approaches build networks where:
>   Nodes represent cited references (co-citation) or citing documents (bibliographic coupling)
>   Edges represent the strength of co-citation or bibliographic coupling between two nodes
>   Clusters represent distinct schools of thought, theoretical frameworks, or research traditions



## General overview


```{r}
devtools::load_all()
library(igraph) # for general network analysis
library(bibliometrix) # to build citation networks
library(visNetwork)
library(plotly)

out_data <- here::here("data", "derived-data")
# load reference list
ref <- read.csv(file.path(out_data, "mte_references_completed.csv"))
ref$Project <- firstup(ref$Project)
# remove references with no DOI
ref <- ref[!is.na(ref$DOI), ]

ref$Relation <- factor(
  ref$Relation,
  levels = c("Proposal", "Top5", "Output"),
  ordered = TRUE
)
ref <- ref[order(ref$Project, ref$Relation, ref$DOI, decreasing = TRUE), ]
# remove duplicates # sum(duplicated(ref))
ref <- ref[!duplicated(ref[, c("Project", "DOI")]), ]

meta <- readxl::read_xlsx(
  here::here("data", "raw-data", "mte", "Projets_FRB-MTECT-OFB.xlsx")
)
# all(meta$Acronyme %in% ref$Project)

# either based on previous script
# M <- readRDS(file.path(out_data, "mte_bibliometrix.rds"))
# dim(M) # 1035, 57
# or from shiny completed references
load(file.path(out_data, "mte_references_oa_shinycomplete.RData"))
# M$CR has short names that can't be linked back to id_oa
# how to get the short name
# dim(M) # 1047, 57
ref$Short <- M$SR[match(ref$DOI, M$DI)]
ref$FAut <- M$AU_CORR[match(ref$DOI, M$DI)]
```



## Co-citation network

```{r}
# get the co-citation network of references
NetMatrix <- bibliometrix::biblioNetwork(
  M,
  analysis = "coupling",
  network = "references",
  n = NULL,
  short = FALSE,
  sep = ";"
)
dim(NetMatrix) # 1035
# table(M$SR %in% dimnames(NetMatrix)[[1]])

# 46722 references with 627 original references

net1 <- igraph::graph_from_adjacency_matrix(
  NetMatrix,
  mode = "undirected",
  diag = FALSE,
  weighted = TRUE
)


# delete non connected literature
# net2 <- delete_vertices(net1, degree(net1) == 0)
net2 <- delete_vertices(net1, components(net1)$membership != 1)

# modularity
mod2 <- cluster_walktrap(net2)
clu2 <- membership(mod2)
sclu2 <- ifelse(clu2 <= (sum(table(clu2) > 1)), clu2, "isolated")
sclu2 <- as.factor(sclu2)
cclu <- colorspace::qualitative_hcl(n = nlevels(sclu2), palette = "Dark 3")
cclu[levels(sclu2) == "isolated"] <- "#808080" # grey
col2 <- cclu[sclu2]


deg2 <- degree(net2)

# V(net1)$size <- sqrt(degree(net1))
# V(netREF)$label <- ref$Short[match(V(net1)$name, ref$Short)]
# V(netREF)$year <- ref$Year[match(V(netREF)$name, ref$DOI)]
# V(netREF)$faut <- ref$FAut[match(V(netREF)$name, ref$DOI)]
# V(netREF)$fulltitle <- ref$Title[match(V(netREF)$name, ref$DOI)]
# V(netREF)$shorttitle <- substr(V(netREF)$fulltitle, 1, 25)
# V(netREF)$title <- paste(
#   V(netREF)$name,
#   V(netREF)$year,
#   V(netREF)$faut,
#   V(netREF)$shorttitle,
#   sep = "<br>"
# )
table(membership(mod2))
V(net2)$group <- membership(mod2)
V(net2)$color <- col2

# check out other layout
visNetwork::visIgraph(
  net2,
  layout = "layout_with_kk",
  idToLabel = FALSE
) |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "group", multiple = T)
  )
```

The network has ```r vcount(net2)``` nodes (one node = one article) and ```r ecount(net2)``` edges (one edge = shared citation between two articles). The connectance is ```r round(edge_density(net2, loops = FALSE),3)``` and the modularity ```r round(modularity(mod2),2)```


```{r}
centR <- data.frame(
  "betweenness" = betweenness(net2),
  "degree" = degree(net2),
  "pagerank" = page_rank(net2)$vector,
  "lab" = V(net2)$name
)

plot_ly(centR) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Reference centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
info <- apply(centR[, -4], 2, round, 5)
score <- apply(scale(sqrt(info)), 1, mean)
info <- as.data.frame(info)
DT::datatable(info[order(score, decreasing = TRUE)[1:50], ])
```


### Per project

```{r}
info <- c()
for (i in unique(ref$Project)) {
  subi <- delete_vertices(
    net2,
    !V(net2)$name %in% ref$Short[ref$Project == i]
  )
  ni <- c("P" = i, "N" = vcount(subi), "C" = edge_density(subi))
  info <- rbind(info, ni)
}
info <- as.data.frame(info)
info$N <- as.numeric(info$N)
info$C <- as.numeric(info$C)
plot_ly(info) |>
  add_markers(
    x = ~N,
    y = ~C,
    text = ~P,
    hoverinfo = "text"
  ) |>
  layout(title = "Project bibliographic coupling") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```