---
title: "Exploration of CESAB groups"
author: Romain Frelat
date: today
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
---


The objective is this document is to explore the diversity and interconnectedness of [FRB CESAB projects](https://www.fondationbiodiversite.fr/en/about-the-foundation/le-cesab/) using network analysis. The exploration consists of two complementary analysis:  

1. project membership network (@sec-member)
2. citation coupling network (@sec-citation)



## 1. Project membership {#sec-member}

```{r}
#| label: data-preprocessing
#| echo: false
#| eval: false

# load reference list
cesab <- readxl::read_xlsx(
  in_data <- here::here("data", "raw-data", "cesab", "CESAB_Members.xlsx")
)
# remove workshop participation
cesab <- cesab[, 1:15]

# simplify affiliation
# table(cesab$affiliation[grepl("univ", tolower(cesab$affiliation))])
cesab$affiliation[grepl("univ", tolower(cesab$affiliation))] <- "University"
cesab$affiliation[cesab$affiliation == "UM"] <- "University"

# Membres fondateurs
mf <- c("CIRAD", "CNRS", "FRB", "IFREMER", "INRAE", "IRD", "MNHN", "OFB")
in_mf <- toupper(cesab$affiliation) %in% mf
cesab$affiliation[in_mf] <- toupper(cesab$affiliation[in_mf])

# Other
cesab$affiliation[!cesab$affiliation %in% c("University", mf)] <- "Other"

# should I keep only actual members
# table(cesab$still_member)

# should I remove guest ?
#fmt: skip
cesab$function_in_group[cesab$function_in_group%in%c("Guesi", "guest")] <- "Guest"
#fmt: skip
cesab$function_in_group[cesab$function_in_group%in%c("Membre")] <- "Member"
#fmt: skip
cesab$function_in_group[cesab$function_in_group%in%c("Post-doc")] <- "Post-Doc"
# table(cesab$function_in_group)

# check names and duplicates
cesab$name <- paste(cesab$lastname, cesab$firstname)

dupname <- duplicated(cesab$name)
dupemail <- duplicated(cesab$emails) & !is.na(cesab$emails)
# table(dupname, dupemail)
# changes of emails : ok
# error <- (dupname & !dupemail)
# ceserr <- cesab[cesab$name %in% cesab$name[error], ]
# ceserr[order(ceserr$name), ]
# errors in name spelling : to be corrected
error <- (dupemail & !dupname)
ceserr <- cesab[cesab$emails %in% cesab$emails[error], ]
ceserr[order(ceserr$name), c("emails", "name")]
cesab$name <- gsub("Guerin Joana", "Guerrin Joana", cesab$name)
cesab$name <- gsub("Rush Adrien", "Rusch Adrien", cesab$name)
cesab$name <- gsub("^Shin Yunne$", "Shin Yunne-Jai", cesab$name)
cesab$name <- gsub("Thackeray Steeve", "Thackeray Stephen", cesab$name)
cesab$name <- gsub("Guillemot Joannès", "Guillemot Joannes", cesab$name)
cesab$name <- gsub("Shatz Bertrand", "Schatz Bertrand", cesab$name)

write.csv(
  cesab,
  here::here("data", "derived-data", "cesab_members.csv"),
  row.names = FALSE
)
```


```{r}
#| label: setup
devtools::load_all()

library(igraph)
library(visNetwork)
library(plotly)
library(bibliometrix) # to build citation networks

# remotes::install_github("patrickbarks/abbrevr")
library(abbrevr)

# load reference list
out_data <- here::here("data", "derived-data")
cesab <- read.csv(
  here::here("data", "derived-data", "cesab_members.csv")
)

# remove guest and stagiaire
cesab <- cesab[!cesab$function_in_group %in% c("Guest", "Stagiaire"), ]

# remove non active members (and keep NA)
cesab <- cesab[!cesab$still_member %in% 0, ]

table(cesab$group, cesab$function_in_group)

nmem <- table(cesab$group)
nproj <- table(table(cesab$name))
# summary(as.numeric(table(cesab$group)))
# # table(duplicated(cesab))
# nrow(cesab) # 1015 membership
```

The FRB-Cesab hosted```r length(nmem)```  projects with ```r length(unique(cesab$name))``` different researchers. Most researchers participated in a single project (N=```r nproj["1"]```), but ```r nproj["2"]``` researchers collaborated to two projects; and ```r sum(nproj[-(1:2)])``` researchers worked on three or more projects.  

On average, a CESAB project includes ```r round(mean(nmem), 1)``` members, with a minimum of ```r  min(nmem)``` for ```r names(which.min(nmem))``` and a maximum of ```r  max(nmem)``` for ```r names(which.max(nmem))```.

### Bipartite membership network

```{r}
netBI <- graph_from_data_frame(
  cesab[, c("name", "group")]
)
V(netBI)$type <- V(netBI)$name %in% cesab$group

V(netBI)$color <- ifelse(V(netBI)$type, "blue", "red")
V(netBI)$shape <- ifelse(V(netBI)$type, "square", "dot")
# V(netBI)$size <- ifelse(V(netBI)$type, 3, 1)

extra <- ifelse(
  V(netBI)$type,
  "",
  cesab$affiliation[match(V(netBI)$name, cesab$name)]
)
V(netBI)$title <- paste(
  V(netBI)$name,
  extra,
  sep = "<br>"
)

# remove members that are not connected to multiple groups
# for visualization purpose only
showBI <- delete_vertices(netBI, degree(netBI) == 1)
# plot(netBI, layout = layout_as_bipartite(netBI))
visNetwork::visIgraph(showBI) |>
  visOptions(
    highlightNearest = TRUE
  )
```

```{r}
splitBI <- bipartite_projection(netBI)
netMEM <- splitBI$proj1
netPR <- splitBI$proj2
```


### Project network

Based on the number of shared members, we can build a network of projects.

```{r}
#netPR2 <- delete_vertices(netPR, degree(netPR) == 1)

# Jaccard transformation to handle uneven nomber of members
# sim <- similarity(netPR, method = "jaccard")
# diag(sim) <- 0
# netPRj <- igraph::graph_from_adjacency_matrix(
#   sim,
#   mode = "undirected",
#   weighted = TRUE
# )
# V(netPRj)$name <- V(netPR)$name

modPR <- cluster_fast_greedy(netPR)
cluPR <- membership(modPR)
nclu <- table(cluPR)
cclu <- colorspace::qualitative_hcl(n = length(nclu), palette = "Dark 3")
cclu[nclu == 1] <- "#808080" # grey
colPR <- cclu[cluPR]
simclu <- ifelse(cluPR <= (sum(nclu > 1)), cluPR, "isolated")
```


The network has ```r vcount(netPR)``` nodes (one node = one project) and ```r ecount(netPR)``` edges (one edge = shared member between two projects). The connectance is ```r round(edge_density(netPR, loops = FALSE),3)``` and the modularity ```r round(modularity(modPR),2)```


```{r}
V(netPR)$group <- simclu
V(netPR)$color <- colPR
# E(netPRj)$width <- E(netPR)$weight
# E(netPRj)$title <- E(netPR)$weight
V(netPR)$title <- V(netPR)$name

visNetwork::visIgraph(
  netPR,
  randomSeed = 54,
  layout = "layout_nicely",
  smooth = TRUE,
  type = "full"
) |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "group", multiple = T)
  )
```


```{r}
# identify cliques
for (i in sort(unique(simclu))) {
  print(paste(
    i,
    paste(V(netPR)$name[simclu == i], collapse = ", "),
    sep = ": "
  ))
}
```

```{r}
centP <- data.frame(
  "degree" = degree(netPR),
  "pagerank" = page_rank(netPR)$vector,
  "betweenness" = betweenness(netPR),
  "lab" = V(netPR)$title,
  "color" = colPR
)

plot_ly(centP) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    marker = list(color = ~color, line = list(color = ~color)),
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Project centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```


```{r}
info <- apply(centP[, 1:3], 2, round, 3)
score <- apply(scale(sqrt(info)), 1, mean)
DT::datatable(info[order(score, decreasing = TRUE), ])
```

### Member centrality

```{r}
E(netMEM)$width <- E(netMEM)$weight
V(netMEM)$size <- log(degree(netMEM))
V(netMEM)$title <- V(netMEM)$name
# visNetwork::visIgraph(netMEM, randomSeed = 54, layout = "layout_nicely") |>
#   visOptions(
#     highlightNearest = TRUE,
#   )
```

```{r}
centR <- data.frame(
  "betweenness" = betweenness(netMEM),
  "degree" = degree(netMEM),
  "pagerank" = page_rank(netMEM)$vector,
  "lab" = V(netMEM)$title
)

plot_ly(centR) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Member centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
info <- apply(centR[, -4], 2, round, 5)
score <- apply(scale(sqrt(info)), 1, mean)
info <- as.data.frame(info)

infoV <- tapply(cesab$group, cesab$name, paste, collapse = ",")
info$project <- infoV[match(row.names(info), names(infoV))]
# info$first_aut <- V(netREF)$faut
# info$year <- V(netREF)$year
# info$short_title <- V(netREF)$shorttitle
DT::datatable(info[order(score, decreasing = TRUE), ])
```


## 2. Citation coupling {#sec-citation}

We estimate the distance among papers produced by the projects by using the bibliographic coupling, i.e. the pairwize overlap in the reference lists. Documents that share many references are likely working on similar topics or using similar methods.  

### Coupling network
```{r}
# missing the association reference <-> project ...
# ref <- read.csv(here::here("data", "raw-data", "cesab", "Zotero_FRB-CESAB.csv"))
# conv <- data.frame(
#   "abb" = ref$Journal.Abbreviation,
#   "ori" = ref$Publication.Title
# )
# conv <- conv[!duplicated(conv), ]

M <- readRDS(file.path(out_data, "cesab_bibliometrix.rds"))

conv <- data.frame(
  "ori" = sort(unique(M$SO))
)
conv$abb <- sapply(firstup(conv$ori), abbrevr::AbbrevTitle, USE.NAMES = FALSE)

# conv <- conv[!duplicated(conv), ]
mconv <- match(tolower(M$SO), tolower(conv$ori))

M$shortjournal <- conv$abb[match(M$SO, conv$ori)]
M$shorttitle <- substr(firstup(M$display_name), 1, 30)
# get the co-citation network of references
NetMatrix <- bibliometrix::biblioNetwork(
  M,
  analysis = "coupling",
  network = "references",
  n = NULL,
  short = FALSE,
  sep = ";"
)
# dim(NetMatrix) # 395
# table(M$SR %in% dimnames(NetMatrix)[[1]])

# 46722 references with 627 original references

net1 <- igraph::graph_from_adjacency_matrix(
  NetMatrix,
  mode = "undirected",
  diag = FALSE,
  weighted = TRUE
)
V(net1)$name <- M$shortname[match(V(net1)$name, M$SR)]

mod1 <- cluster_fast_greedy(net1)


# delete non connected literature
# net2 <- delete_vertices(net1, degree(net1) == 0)
net2 <- delete_vertices(net1, components(net1)$membership != 1)

# modularity
mod2 <- cluster_fast_greedy(net2)
clu2 <- membership(mod2)
sclu2 <- ifelse(clu2 <= (sum(table(clu2) > 1)), clu2, "isolated")
sclu2 <- as.factor(sclu2)
cclu <- colorspace::qualitative_hcl(n = nlevels(sclu2), palette = "Dark 3")
cclu[levels(sclu2) == "isolated"] <- "#808080" # grey
col2 <- cclu[sclu2]

deg2 <- degree(net2)
```

The network has ```r vcount(net1)``` nodes (one node = one article), and ```r scales::comma(ecount(net1))``` edges (one edge = shared references between two articles). The connectance is ```r round(edge_density(netPR, loops = FALSE),3)``` and the modularity ```r round(modularity(modPR),2)```


For visual representation, we only keep the largest connected community of articles (we removed the disconnected articles).
```{r}
# V(net1)$size <- sqrt(degree(net1))
# V(netREF)$label <- ref$Short[match(V(net1)$name, ref$Short)]
# V(netREF)$year <- ref$Year[match(V(netREF)$name, ref$DOI)]
# V(netREF)$faut <- ref$FAut[match(V(netREF)$name, ref$DOI)]
# V(netREF)$fulltitle <- ref$Title[match(V(netREF)$name, ref$DOI)]
# V(netREF)$shorttitle <- substr(V(netREF)$fulltitle, 1, 25)

V(net2)$title <- paste(
  V(net2)$name,
  M$shortjournal[match(V(net2)$name, M$shortname)],
  M$shorttitle[match(V(net2)$name, M$shortname)],
  sep = "<br>"
)
V(net2)$group <- membership(mod2)
V(net2)$color <- col2

# check out other layout
visNetwork::visIgraph(
  net2,
  layout = "layout_with_kk",
  idToLabel = FALSE
) |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "group", multiple = T)
  )
```


### Article centrality

```{r}
centR <- data.frame(
  "betweenness" = betweenness(net2),
  "degree" = degree(net2),
  "pagerank" = page_rank(net2)$vector,
  "color" = V(net2)$color,
  "lab" = V(net2)$title
)

plot_ly(centR) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    marker = list(color = ~color, line = list(color = ~color)),
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Reference centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
info <- apply(centR[, 1:3], 2, round, 5)
score <- apply(scale(sqrt(info)), 1, mean)
info <- as.data.frame(info)

info$shortjournal <- M$shortjournal[match(row.names(info), M$shortname)]
info$shorttitle <- M$shorttitle[match(row.names(info), M$shortname)]
DT::datatable(info[order(score, decreasing = TRUE)[1:50], ])
```

## Conclusion

- missing year of project for a temporal analysis
- missing the link between project and references (how to export it from Zenodo?)
