---
title: "Exploration of CESAB groups"
author: Romain Frelat
date: today
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
---


The objective is this document is to explore the diversity and interconnectedness of [FRB CESAB projects](https://www.fondationbiodiversite.fr/en/about-the-foundation/le-cesab/) using network analysis. The exploration consists of two complementary analysis:   

1. [Project membership network](#project-membership) based on who is collaborating in the projects   
2. [Citation coupling network](#citation-coupling)  based on the scientific articles published by projects



## 1. Project membership

```{r}
#| label: setup
#| echo: false
devtools::load_all()

library(igraph)
library(visNetwork)
library(plotly)
library(bibliometrix) # to build citation networks

# remotes::install_github("patrickbarks/abbrevr")
library(abbrevr)

# load reference list
out_data <- here::here("data", "derived-data")
cesab <- read.csv(
  here::here("data", "derived-data", "cesab_members.csv")
)

# remove guest and stagiaire
cesab <- cesab[!cesab$function_in_group %in% c("Guest", "Stagiaire"), ]

# remove non active members (and keep NA)
cesab <- cesab[!cesab$still_member %in% 0, ]

# remove duplicates (mostly for Pelagic 1 and 2)
cesab <- cesab[!duplicated(cesab[, c("group", "name")]), ]

infoV <- tapply(cesab$group, cesab$name, paste, collapse = "<br>")

nmem <- table(cesab$group)
nproj <- table(table(cesab$name))
# summary(as.numeric(table(cesab$group)))
# # table(duplicated(cesab))
# nrow(cesab) # 1015 membership

# citation dataset
M <- readRDS(file.path(out_data, "cesab_bibliometrix.rds"))

conv <- data.frame(
  "ori" = sort(unique(M$SO))
)
conv$abb <- sapply(firstup(conv$ori), abbrevr::AbbrevTitle, USE.NAMES = FALSE)
# conv <- conv[!duplicated(conv), ]
mconv <- match(tolower(M$SO), tolower(conv$ori))

M$shortjournal <- conv$abb[match(M$SO, conv$ori)]
M$shorttitle <- substr(firstup(M$display_name), 1, 30)
M$breaktitle <- breaks_string(firstup(M$display_name))
M$linkDOI <- paste0(
  "<a href=\'",
  paste0("https://www.doi.org/", M$DI),
  "\'>",
  M$DI,
  "</a>"
)

M$project <- gsub("PPR Oceans", "PPR Océans", M$project)
# keep only references in CESAB group (or in Core)
# remove Desybel
M <- M[!M$project %in% "Desybel", ]
```



The FRB-Cesab hosted```r length(nmem)```  projects between ```r min(cesab$start_year)``` and ```r max(cesab$start_year)``` with ```r length(unique(cesab$name))``` different researchers. On average, a CESAB project includes ```r round(mean(nmem), 1)``` members, with a minimum of ```r  min(nmem)``` for ```r names(which.min(nmem))``` and a maximum of ```r  max(nmem)``` for ```r names(which.max(nmem))``` (@tbl-mempro). Most researchers participated in a single project (N=```r nproj["1"]```), but ```r nproj["2"]``` researchers collaborated in two projects; and ```r sum(nproj[-(1:2)])``` researchers worked on three or more projects.  



```{r}
#| label: tbl-mempro
#| tbl-cap: "Number of members, principal investigators (PI), and post-doc per CESAB project."
#| 
table(cesab$group, cesab$function_in_group)
```

### Bipartite membership network

```{r}
#| label: fig-bimem
#| fig-cap: "Simplified bipartite network containing all projects (nodes in blue square) and the members that participated in multiple projects (nodes in red circle)."
#| 
netBI <- graph_from_data_frame(
  cesab[, c("name", "group")]
)
V(netBI)$type <- V(netBI)$name %in% cesab$group

V(netBI)$color <- ifelse(V(netBI)$type, "blue", "red")
V(netBI)$shape <- ifelse(V(netBI)$type, "square", "dot")
# V(netBI)$size <- ifelse(V(netBI)$type, 3, 1)

extra <- ifelse(
  V(netBI)$type,
  "",
  infoV[match(V(netBI)$name, names(infoV))]
)
V(netBI)$title <- paste(
  V(netBI)$name,
  extra,
  sep = "<br>"
)

# remove members that are not connected to multiple groups
# for visualization purpose only
showBI <- delete_vertices(netBI, degree(netBI) == 1)

visNetwork::visIgraph(
  showBI,
  randomSeed = 25,
  layout = "layout_with_fr", #layout_with_kk, layout_nicely
  smooth = TRUE
) |>
  visOptions(
    highlightNearest = TRUE
  )
```

The bipartite network offer a complete visualization of the projects (blue square) and their members (red circles) (@fig-bimem). However, the complexity of handling two types of nodes (project and member) prevent further in-depth analysis. The bipartite network can be projected into (1) [a project network](#project-network), and (2) [a member network](#member-network).  

```{r}
#| echo: false
splitBI <- bipartite_projection(netBI)
netMEM <- splitBI$proj1
netPR <- splitBI$proj2
```


### Project network

The project network is made of nodes representing projects and edges representing the number of shared members.

```{r}
# Jaccard transformation to handle uneven nomber of members
# sim <- similarity(netPR, method = "jaccard")
# diag(sim) <- 0
# netPRj <- igraph::graph_from_adjacency_matrix(
#   sim,
#   mode = "undirected",
#   weighted = TRUE
# )
# V(netPRj)$name <- V(netPR)$name

modPR <- cluster_fast_greedy(netPR)
cluPR <- membership(modPR)
nclu <- table(cluPR)
simclu <- ifelse(cluPR <= (sum(nclu > 1)), cluPR, "isolated")
# add0 <- cluPR < 10 & simclu != "isolated"
# simclu[add0] <- paste0("0", simclu[add0])
simclu <- as.factor(simclu)
cclu <- colorspace::qualitative_hcl(n = nlevels(simclu) - 1, palette = "Dark 3")
cclu <- c(cclu, "#808080") # grey
colPR <- cclu[simclu]
```


The network has ```r vcount(netPR)``` nodes (one node = one project) and ```r ecount(netPR)``` edges (one edge = shared member between two projects). The connectance is ```r round(edge_density(netPR, loops = FALSE),3)``` (very low) and the modularity is ```r round(modularity(modPR),2)``` (moderately high). It means that there are few connections among projects but they are organized in highly connected sub-groups (called modules).

#### Modules

```{r}
#| label: fig-netmempro
#| fig-cap: "Project network of shared membership. The width of the edges corresponds to the number of members shared between projects. The different modules of projects are highlighted in different colors."
#| 

V(netPR)$group <- simclu
V(netPR)$color <- colPR
E(netPR)$width <- E(netPR)$weight
E(netPR)$title <- E(netPR)$weight
V(netPR)$year <- cesab$start_year[match(V(netPR)$name, cesab$group)]
V(netPR)$title <- paste(V(netPR)$name, V(netPR)$year, sep = "<br>")

visNetwork::visIgraph(
  netPR,
  randomSeed = 25,
  layout = "layout_with_fr",
  smooth = TRUE,
  type = "full"
) |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "group", multiple = T)
  )
```


```{r}
# echo: false
# identify cliques
for (i in sort(unique(simclu))) {
  print(paste(
    i,
    paste(V(netPR)$name[simclu == i], collapse = ", "),
    sep = ": "
  ))
}
```

#### Temporal dynamic

```{r}
#| label: fig-netyrmempro
#| fig-cap: "Project network of shared membership visualized per starting year (in x-axis) and per module (in y-axis)."
#| 

# x-axis by year
vyr <- V(netPR)$year
xnod <- (vyr - min(vyr)) / (max(vyr) - min(vyr)) * 2 - 1

# take the membership as y-axis
ymod <- as.numeric(as.factor(simclu))
ynod <- (ymod / max(ymod)) * 2 - 1
# or take the first MDS dimension
# ynod <- layout_with_mds(netMEM)[, 1]

# coordinates
coo <- cbind(jitter(xnod, 1.5), jitter(ynod, 3))

# plot(netMEM, layout = coo)

visNetwork::visIgraph(
  netPR,
  layout = 'layout.norm',
  layoutMatrix = coo,
  randomSeed = 54,
  type = "full"
) |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "year", multiple = T)
  )
```

#### Project centrality

To quantify the project centrality, i.e. how project connects different parts of the network, we examine three indices:  

- Degree: number of connections to other projects;  
- Google PageRank: influence based on the number and the importance of connected project;  
- Betweenness: number of shortest paths going through a project (calculated for the largest connected graph)

```{r}
#| label: fig-cenpro
#| fig-cap: "Project centrality based on shared members. The size of the dot corresponds to the degree (number of connections), its colors corresponds to the modules. "
#| 
centP <- data.frame(
  "degree" = degree(netPR),
  "pagerank" = page_rank(netPR)$vector,
  "betweenness" = betweenness(netPR),
  "lab" = V(netPR)$title,
  "color" = colPR
)

plot_ly(centP) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    marker = list(color = ~color, line = list(color = ~color)),
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Project centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```


```{r}
#| label: tbl-cenpro
#| tbl-cap: "Project centrality based on shared members. The degree is the number of connections to other projects, the Google PageRank measures the influence based on the number and the importance of connected project and the betweenness is the number of shortest paths going through a project."
info <- apply(centP[, 1:3], 2, round, 3)
score <- apply(scale(sqrt(info)), 1, mean)
DT::datatable(info[order(score, decreasing = TRUE), ])
```


### Member network

The member network is made of nodes representing members and edges representing project collaboration. In other words, all members of a project are connected to each other. Because of the high number of nodes (```r vcount(netMEM)``` members), it is not informative to plot the network, but it provides relevant information about members' centrality.


#### Member centrality

The member centrality quantifies how individuals connect different parts of the network. As previously, it is quantified by three indicators: degree, Google PageRank, and Betweenness.   

```{r}
#| label: fig-cenmem
#| fig-cap: "Member centrality based on project collaboration. The size of the dot corresponds to the degree (number of connections)."

# Visualization
E(netMEM)$width <- E(netMEM)$weight
V(netMEM)$size <- log(degree(netMEM))
V(netMEM)$title <- paste(
  V(netMEM)$name,
  infoV[match(V(netMEM)$name, names(infoV))],
  sep = "<br>"
)

# visNetwork::visIgraph(netMEM, randomSeed = 54, layout = "layout_with_fr") |>
#   visOptions(
#     highlightNearest = TRUE,
#   )

# Centrality
centR <- data.frame(
  "betweenness" = betweenness(netMEM),
  "degree" = degree(netMEM),
  "pagerank" = page_rank(netMEM)$vector,
  "lab" = V(netMEM)$title
)

plot_ly(centR) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Member centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
#| label: tbl-cenmem
#| tbl-cap: "Member's centrality based on project collaboration. The degree is the number of collaborations, the Google PageRank measures the influence based on the number and the importance of collaborations and the betweenness is the number of shortest paths going through a member."
info <- apply(centR[, -4], 2, round, 5)
score <- apply(scale(sqrt(info)), 1, mean)
info <- as.data.frame(info)

infoV <- tapply(cesab$group, cesab$name, paste, collapse = ", ")
info$project <- infoV[match(row.names(info), names(infoV))]
# info$first_aut <- V(netREF)$faut
# info$year <- V(netREF)$year
# info$short_title <- V(netREF)$shorttitle
DT::datatable(info[order(score, decreasing = TRUE), ])
```


## 2. Citation coupling 

Another way to measure the closeness or distance between scientific projects is to use the scientific articles published from the project activities.

The CESAB projects published ```r nrow(M)```  scientific articles between ```r min(M$PY)``` and ```r max(M$PY)```. We estimate the distance among articles by using the bibliographic coupling, i.e. the pairwize overlap in the reference lists. Documents that share many references are likely working on similar topics or using similar methods.  

### Coupling network

```{r}
# get the co-citation network of references
NetMatrix <- bibliometrix::biblioNetwork(
  M,
  analysis = "coupling",
  network = "references",
  n = NULL,
  short = FALSE,
  sep = ";"
)
# dim(NetMatrix) # 406

netCC <- igraph::graph_from_adjacency_matrix(
  NetMatrix,
  mode = "undirected",
  diag = FALSE,
  weighted = TRUE
)
V(netCC)$name <- M$shortname[match(V(netCC)$name, M$SR)]

modCC <- cluster_louvain(netCC)

# delete non connected literature
# not needed anymore
# netCC <- delete_vertices(netCC, degree(netCC) == 0)
netCC2 <- delete_vertices(netCC, components(netCC)$membership != 1)
# table(components(netCC)$membership)

# modularity
modCC2 <- cluster_louvain(netCC2)
cluCC2 <- membership(modCC2)
# group isolated modules
# scluCC <- ifelse(cluCC <= (sum(table(cluCC2) > 1)), cluCC, "isolated")
# add 0 for nicer sorting
add0 <- cluCC2 < 10
cluCC2[add0] <- paste0("0", cluCC2[add0])
cluCC2 <- as.factor(cluCC2)
# attribute colors
palCC2 <- colorspace::qualitative_hcl(n = nlevels(cluCC2), palette = "Dark 3")
colCC2 <- palCC2[cluCC2]
```

The network has ```r vcount(netCC)``` nodes (one node = one article), and ```r scales::comma(ecount(netCC), big.mark = " ")``` edges (one edge = shared references between two articles). The connectance is ```r round(edge_density(netCC, loops = FALSE),3)``` and the modularity ```r round(modularity(modCC),2)```.   

For visual representation, we only keep the largest connected community of articles (we removed the disconnected articles, N=```r sum(components(netCC)$membership != 1)```).

```{r}
#| label: fig-ref
#| fig-cap: "Largest connected community of the coupling network. Nodes are articles and edges represent the shared cited references. The colors represent the modules of the coupling network."

V(netCC2)$title <- paste(
  V(netCC2)$name,
  M$shortjournal[match(V(netCC2)$name, M$shortname)],
  M$breaktitle[match(V(netCC2)$name, M$shortname)],
  M$linkDOI[match(V(netCC2)$name, M$shortname)],
  M$project[match(V(netCC2)$name, M$shortname)],
  sep = "<br>"
)
V(netCC2)$group <- cluCC2
V(netCC2)$color <- colCC2

# check out other layout
visNetwork::visIgraph(
  netCC2,
  layout = "layout_with_fr",
  idToLabel = FALSE,
  randomSeed = 25,
  # smooth = TRUE,
  type = "full"
) |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "group", multiple = T)
  )
```


#### Article centrality

```{r}
#| label: fig-cenref
#| fig-cap: "Article centrality based on citation coupling. The size of the dot corresponds to the degree (number of connections), its colors corresponds to the modules. "
centR2 <- data.frame(
  "betweenness" = betweenness(netCC2),
  "degree" = degree(netCC2),
  "pagerank" = page_rank(netCC2)$vector,
  "color" = V(netCC2)$color,
  "lab" = V(netCC2)$title
)

plot_ly(centR2) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    marker = list(color = ~color, line = list(color = ~color)),
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Article centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
#| label: tbl-cenref
#| tbl-cap: "Article centrality based on citation coupling. The degree is the number of connections to other articles, the Google PageRank measures the influence based on the number and the importance of connected references and the betweenness is the number of shortest paths going through a reference."

# compute centrality on all references
centR <- data.frame(
  "betweenness" = round(betweenness(netCC), 1),
  "degree" = round(degree(netCC), 1),
  "pagerank" = round(page_rank(netCC)$vector, 5)
)

score <- apply(scale(sqrt(centR)), 1, mean)
centR <- as.data.frame(centR)

centR$journal <- M$shortjournal[match(row.names(centR), M$shortname)]
# centR$title <- M$breaktitle[match(row.names(centR), M$shortname)]
centR$project <- M$project[match(row.names(centR), M$shortname)]
centR$DOI <- M$linkDOI[match(row.names(centR), M$shortname)]
DT::datatable(centR[order(score, decreasing = TRUE), ], escape = FALSE)
```


#### Centrality per project

Each article is attached to one or multiple projects. We can use this link to calculate the average article centrality per project.

```{r}
#| label: fig-cenrefpro
#| fig-cap: "Average article centrality per project. The size of the dot corresponds to the number of articles per project. "
#| 

pp <- strsplit(centR$project, ", ")
# data frame per reference
infoR <- data.frame(
  "project" = unlist(pp),
  "betweenness" = rep(centR$betweenness, sapply(pp, length)),
  "degree" = rep(centR$degree, sapply(pp, length)),
  "pagerank" = rep(centR$pagerank, sapply(pp, length))
)

# statistics per project
infoP <- data.frame(
  "project" = sort(unique(infoR$project)),
  "betweenness" = round(tapply(infoR$betweenness, infoR$project, mean), 1),
  "betweenness_sd" = tapply(infoR$betweenness, infoR$project, sd),
  "degree" = round(tapply(infoR$degree, infoR$project, mean), 1),
  "pagerank" = round(tapply(infoR$pagerank, infoR$project, mean), 5),
  "pagerank_sd" = tapply(infoR$pagerank, infoR$project, sd),
  "N" = as.numeric(table(infoR$project))
)
# infoP$project[!infoP$project %in% cesab$group]
infoP$year <- cesab$start_year[match(infoP$project, cesab$group)]
infoP$lab <- paste(
  infoP$project,
  ifelse(is.na(infoP$year), "", infoP$year),
  "<br>N ref =",
  infoP$N
)

plot_ly(
  infoP,
  x = ~betweenness,
  y = ~pagerank,
  # color = ~project,
  size = ~N,
  # marker = list(color = ~project, line = list(color = ~project)),
  text = ~lab,
  hoverinfo = "text",
  type = "scatter",
  mode = "markers",
  # error_x = ~ list(array = betweenness_sd),
  # error_y = ~ list(array = pagerank_sd),
  showlegend = FALSE
) |>
  layout(title = "Average article centrality per project") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
#| label: tbl-cenrefpro
#| tbl-cap: "Average reference centrality per project. The degree is the number of shared references, the Google PageRank measures the influence based on the number and the importance of connected references and the betweenness is the number of shortest paths going through a reference."
#| 
info <- infoP[, c("project", "year", "N", "degree", "pagerank", "betweenness")]
score <- apply(
  scale(sqrt(infoP[, c("degree", "pagerank", "betweenness")])),
  1,
  mean
)

DT::datatable(
  info[order(score, decreasing = TRUE), ],
  escape = FALSE,
  rownames = FALSE
)
```


## Conclusion

- already some cool networks analysis that provide a good overview of CESAB's projects  

*Exploration to be completed...*  


