---
title: "Exploration of project citations"
author: Romain Frelat
date: today
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
---


The objective is to explore the associations among [FRB-MTE-OFB projects](https://www.fondationbiodiversite.fr/la-frb-en-action/programmes-et-projets/impacts-sur-la-biodiversite-terrestre-dans-lanthropocene/) based on shared references. Shared references can reveal common conceptual frameworks, methodological approaches, or research topics across projects.

Based on references, we:  

- cluster projects based on shared references 
- identify papers that act as bridges between multiple projects (i.e. *hubs*)

The analysis is based on a bipartite network linking projects and references, which is subsequently projected into project and reference networks. This document is mostly an exploration of (bipartite) network analysis methodology.  


## General overview

```{r}
devtools::load_all()
library(igraph) # network analysis
library(visNetwork) # interactive network
library(plotly) # interactive plot

# load reference list
ref <- read.csv(
  here::here("data", "derived-data", "mte_references_completed.csv")
)
ref$Project <- firstup(ref$Project)
# remove references with no DOI
ref <- ref[!is.na(ref$DOI), ]

ref$Relation <- factor(
  ref$Relation,
  levels = c("Proposal", "Top5", "Output"),
  ordered = TRUE
)
ref <- ref[order(ref$Project, ref$Relation, ref$DOI, decreasing = TRUE), ]
# remove duplicates # sum(duplicated(ref))
ref <- ref[!duplicated(ref[, c("Project", "DOI")]), ]

# load informations on projects
meta <- readxl::read_xlsx(
  here::here("data", "raw-data", "mte", "Projets_FRB-MTECT-OFB.xlsx")
)

ndoi <- table(table(ref$DOI))

# complete the references with author names
load(
  here::here("data", "derived-data", "mte_references_oa_shinycomplete.RData")
)
# M <- readRDS(here::here("data", "derived-data", "mte_bibliometrix.rds"))
# table(ref$DOI %in% M$DI)
ref$Short <- M$SR[match(ref$DOI, M$DI)]
ref$FAut <- M$AU_CORR[match(ref$DOI, M$DI)]
```

```{r}
# table(ref$Year, ref$Project)
table(ref$Project, ref$Relation)

# between 11 and 116 references per project
# summary(as.numeric(table(ref$Project)))
n_ref <- data.frame(table(ref$Project))
n_ref$Type <- meta$Type[match(n_ref$Var1, meta$Acronyme)]
tapply(n_ref$Freq, n_ref$Type, summary)
```


Each project cited between ```r min(n_ref$Freq)``` and ```r max(n_ref$Freq)``` references in their proposal. These references come from the main text (`Proposal`), the top articles per member (`Top5`) or listed as `Output` in the final report. There are strong differences in the proposal format depending on the project type. `Synthese` projects have the most complete proposal compare to `Synergie` and `Revue systematique`.  

In total, there are ```r nrow(ref)``` references listed by the ```r nrow(n_ref)``` projects. Most publications are cited by a single project (N=```r ndoi["1"]```), but ```r sum(ndoi[-1])``` publications are cited by at least two projects.


## Bipartite network

We can build a bi-partite network in which references connect projects.

```{r}
netBI <- graph_from_data_frame(
  ref[, c("DOI", "Project")]
)
V(netBI)$type <- V(netBI)$name %in% ref$Project
# V(netBI)$size <- 1
# modBI <- cluster_walktrap(netBI)
```

The bipartite network has ```r vcount(netBI)``` nodes (project or reference), ```r ecount(netBI)``` edges, and a connectance of ```r round(edge_density(netBI, loops = FALSE),3)```. 

```{r}
V(netBI)$color <- ifelse(V(netBI)$type, "red", "blue")
V(netBI)$shape <- ifelse(V(netBI)$type, "square", "dot")

V(netBI)$label <- ifelse(V(netBI)$type, V(netBI)$name, "")
# V(netBI)$size <- degree(netBI)
extra <- ifelse(
  V(netBI)$type,
  paste(
    meta$Type[match(V(netBI)$name, meta$Acronyme)],
    meta$Year[match(V(netBI)$name, meta$Acronyme)],
    sep = "<br>"
  ),
  ref$Short[match(V(netBI)$name, ref$DOI)]
)

V(netBI)$title <- paste(
  V(netBI)$name,
  extra,
  sep = "<br>"
)

showBI <- delete_vertices(netBI, degree(netBI) == 1)

# plot(netBI, layout = layout_as_bipartite(netBI))
visNetwork::visIgraph(showBI, idToLabel = FALSE, smooth = TRUE) |>
  visLegend(
    addNodes = list(
      list(
        label = "Project",
        shape = "square",
        color = "red"
      ),
      list(
        label = "Reference",
        shape = "dot",
        color = "blue"
      )
    ),
    useGroups = FALSE
  ) |>
  visOptions(
    highlightNearest = TRUE,
  )
```


To facilitate interpretation, the bipartite network is projected into two complementary networks:   

- a project network to explore similarities among projects
- a reference network to identify the most central references


```{r}
splitBI <- bipartite_projection(netBI)
netREF <- splitBI$proj1
netPR <- splitBI$proj2
```

### Project network

Based on the number of shared references, we can build a network of projects.

```{r}
# manually count references
# group_mem <- table(mem$group, mem$name)
# member_overlap <- tcrossprod(group_mem)
# diag(member_overlap) <- 0
# netPR <- igraph::graph_from_adjacency_matrix(
#   member_overlap,
#   mode = "undirected",
#   weighted = TRUE
# )

# weigthed by number of references
# sim <- similarity(netPR, method = "jaccard")
# diag(sim) <- 0
# netPR <- igraph::graph_from_adjacency_matrix(
#   sim,
#   mode = "undirected",
#   weighted = TRUE
# )

modPR <- cluster_walktrap(netPR)
cluPR <- membership(modPR)
nclu <- table(cluPR)
cclu <- colorspace::qualitative_hcl(n = length(nclu), palette = "Dark 3")
cclu[nclu == 1] <- "#808080" # grey
colPR <- cclu[cluPR]
simclu <- ifelse(cluPR <= (sum(nclu > 1)), cluPR, "isolated")
```



The network has ```r vcount(netPR)``` nodes (one node = one project) and ```r ecount(netPR)``` edges (one edge = shared reference between two projects). The connectance is ```r round(edge_density(netPR, loops = FALSE),3)``` and the modularity ```r round(modularity(modPR),2)```


```{r}
V(netPR)$type <- as.factor(meta$Type[match(V(netPR)$name, meta$Acronyme)])
V(netPR)$shape <- c("triangle", "square", "dot")[V(netPR)$type]
V(netPR)$year <- meta$Year[match(V(netPR)$name, meta$Acronyme)]
V(netPR)$group <- simclu
V(netPR)$color <- colPR
E(netPR)$width <- E(netPR)$weight
E(netPR)$title <- E(netPR)$weight
V(netPR)$title <- paste(
  V(netPR)$name,
  V(netPR)$type,
  V(netPR)$year,
  sep = "<br>"
)
visNetwork::visIgraph(
  netPR,
  randomSeed = 54,
  layout = "layout_nicely",
  smooth = TRUE,
  type = "full"
) |>
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(variable = "group", multiple = T)
  ) |>
  visLegend(
    addNodes = list(
      list(
        label = "Revue",
        shape = "triangle",
        color = "grey"
      ),
      list(
        label = "Synergie",
        shape = "square",
        color = "grey"
      ),
      list(
        label = "Synthese",
        shape = "dot",
        color = "grey"
      )
    ),
    useGroups = FALSE
  ) |>
  visPhysics(stabilization = TRUE)
```

```{r}
# identify cliques
for (i in sort(unique(simclu))) {
  print(paste(
    i,
    paste(V(netPR)$name[simclu == i], collapse = ", "),
    sep = ": "
  ))
}
```


## Reference network

The reference network links publications that are cited by the same projects. Central references are therefore publications that contribute to connecting different parts of the project landscape.

We examine three centrality indices:


- Betweenness: number of shortest paths going through a reference
- Degree: number of connections to other references;
- Google PageRank: influence based on the number and the importance of connected references.


```{r}
# show the weight
E(netREF)$width <- E(netREF)$weight

V(netREF)$size <- log(degree(netREF))

V(netREF)$label <- ref$Short[match(V(netREF)$name, ref$DOI)]
V(netREF)$year <- ref$Year[match(V(netREF)$name, ref$DOI)]
V(netREF)$faut <- ref$FAut[match(V(netREF)$name, ref$DOI)]
V(netREF)$fulltitle <- ref$Title[match(V(netREF)$name, ref$DOI)]
V(netREF)$shorttitle <- substr(V(netREF)$fulltitle, 1, 25)

V(netREF)$title <- paste(
  V(netREF)$name,
  V(netREF)$year,
  V(netREF)$faut,
  V(netREF)$shorttitle,
  sep = "<br>"
)

# plot(netREF)
# visNetwork::visIgraph(netREF, randomSeed = 54, layout = "layout_nicely") |>
#   visOptions(
#     highlightNearest = TRUE,
#   )
```




```{r}
centR <- data.frame(
  "betweenness" = betweenness(netREF),
  "degree" = degree(netREF),
  "pagerank" = page_rank(netREF)$vector,
  "lab" = V(netREF)$title
)

plot_ly(centR) |>
  add_markers(
    x = ~betweenness,
    y = ~pagerank,
    size = ~degree,
    text = ~lab,
    hoverinfo = "text"
  ) |>
  layout(title = "Reference centrality") |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

```{r}
info <- apply(centR[, -4], 2, round, 5)
score <- apply(scale(sqrt(info)), 1, mean)
info <- as.data.frame(info)
info$first_aut <- V(netREF)$faut
info$year <- V(netREF)$year
info$short_title <- V(netREF)$shorttitle
DT::datatable(info[order(score, decreasing = TRUE)[1:50], ])
```

## Conclusion

We identified 5 sub-groups of projects based on shared literature and a handful of the most influential references among the FRB-MTE-OFB projects.   

Yet, these results are drawn only from the *raw* reference list, ignoring the relation between the articles, which could be explored by a co-citation network. To further estimate the breadth and interdisciplinarity of research projects, we need integrate the citations among a larger pool of articles.